\documentclass{article}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{array}
\usepackage{verbatim}

\geometry{
total={160mm,257mm},
left=20mm,
top=15mm,
bottom=20mm
}

\begin{document}
    \begin{center}
        {\huge Brute Force}

        Author: Taylor Wong
    \end{center}

    In this video we're going to be going over one of the simplest and yet one of the most complex problem solving paradigms: Brute Force. In short the brute force approach is about trying possible "potential solutions" until one is proven to work. This is an extremely powerful approach since it can be used to solve any problem where we can verify a solution quickly. These problems belong to a group of problems called NP which includes the vast majority of problems in competitive programming. So if it's so powerful why don't we use it for every problem? The number of potential solutions usually grows exponentially or faster. For example, consider the sorting problem. We're given an array of $n$ integers and have to return an array with the same integers but in sorted order. Our potential solutions will be all permutations of the original integers. It takes $O(n)$ time to verify if a potential solution is correct (simply scan through the array and if any elements are out of order return false). The number of permutations however is $n!$ so our overall runtime (in the worst case) will be $O(n!)$ which is obviously much worse than our usual $O(n\log n)$. Ok so if brute force approaches tend to be so slow why would we use them? While it's true that the majority of problems you'll come across in competitive programming will have faster solutions than brute force there are three types of problems where you might want to use brute force.
    \begin{enumerate}
        \item Problems where the number of potential solutions doesn't grow quickly.
        \item Problems where the maximum input size is very small.
        \item Problems that cannot be solved in any other way (NP-Complete)
    \end{enumerate}

\end{document}
